---
title: "project2"
author: "Nancy Truong"
date: "2025-04-15"
output: html_document
---
# 1. Low plasma β-carotene and vitamin use
## (a)
```{r}
library(tidyverse); theme_set(theme_bw() + theme(text = element_text(size = 18)))
getwd()  # Check current working directory
list.files("Data")  # Check if 'carotene.xlsx' is inside the "Data" folder
install.packages("readxl")  # Only run once if not installed
library(readxl)
library(dplyr)  # Load dplyr for select()
# Lecture 3 - Elasticity - multiple linear regression

library(car)
library(rstatix)
library(GGally)
df <- read_excel("Data/carotene.xlsx")
head(df)
```
0.42 μmol/l (micromoles per litre)
= ? ng/ml  (nanograms per millilitre)

β-carotene: C40H56
```{r}
plasmab_molar_mass <- 40*12.010788+56*1.0079477
alpha <- 0.42*plasmab_molar_mass
# 225.4882
summary(df)
# 3rd Qu.: 230.0  
```

## (b)
Create a new variable, lowplasma_01, that is 1 when betaplasma < a ng/ml, and 0 otherwise.

```{r}
#?ifelse
#plasmab$lowplasma_01 <- ifelse(plasmab$betaplasma < alpha, 1, 0)
glimpse(df)

df |> mutate(
lowplasma_01 = as.numeric(betaplasma < alpha),
lowplasma_hl = factor(lowplasma_01,
levels = c(0, 1),
labels = c("high", "low"))) -> df
```
You can then plot with, e.g., aes(..., y = lowplasma_01) and get 0/1 on the y-axis, while using aes(..., color = lowplasma_hl) to get separate colours for the high and
low categories.
```{r}
ggplot(df, aes(age, lowplasma_01, color = lowplasma_hl)) +
  geom_point() +
  geom_smooth() +
  xlab("") +
  ylab("Low concentration of plasma β-carotene") +
  labs(title = "Low plasma β-carotene(=1) or Not low β-carotene (=0) vs x") 
```
In the report, present the number of observations that have "low" and "high" plasma β-carotene concentrations, respectively, together with the proportion with "low" concentration, as a percentage.
```{r}
lowplasma <- as.data.frame(
  count(df, lowplasma_01))
lowplasma$percentage <- round(lowplasma$n*100/sum(lowplasma$n),1)
lowplasma
```


## (c)
Start by examining the relationship between low plasma β-carotene and vitamin use by counting the number of observations for all combinations, using count, and then transform the result into a wide format with pivot_wider(), resulting in a 3 × 2-table
```{r}
plasma_vituse <- df |> count(vituse, lowplasma_hl) |>
pivot_wider(id_cols = vituse,
names_from = lowplasma_hl,
values_from = n)
plasma_vituse
```


Calculate the probabilities and the corresponding odds for having low plasma β-carotene for each of the three vitamin use categories, and add them to the table (mutate()).
```{r}
plasma_vituse
# Prob. for having low plasma B
# Correspnding odds
# odds = Pr(success) / Pr(failure)
plasma_vituse <- plasma_vituse |>
  mutate(
    prob_low = low*100/(low + high),
    odds = prob_low / (100-prob_low)
  )
plasma_vituse
```
Also calculate the odds ratios for each of the categories, compared to a suitable reference category,
and add them to the table, Table.1(c).

```{r}
mutate(df,
       vituse = factor(vituse,
                         levels = c(1, 2, 3),
                         labels = c("often", "notoften", "no"))) -> df
# odd ratios = OR = oddsj/ odd ref
plasma_vituse <- plasma_vituse |>
  mutate(
   OR = odds / plasma_vituse$odds[1]
  )
plasma_vituse



```

## (d)
Fit a logistic regression model, Model.1(d), for lowplasma_01 or lowplasma_hl with vituse as explanatory variable.

Present the β-estimates, the e^β -estimates and their profile likelihood based 95 % confidence intervals.
-The confint()-function calculates profile likelihood based confidence intervals for the parameters.
```{r}
vituse_glm <- glm(lowplasma_01 ~ vituse, family = "binomial", data = df)
vituse_glm
vituse_sum <- summary(vituse_glm)
vituse_sum
vituse_sum$coefficients
bhat <- vituse_glm$coefficients
ci.beta <- confint(vituse_glm)
or = exp(bhat)
ci.or <- exp(ci.beta)
cbind('beta'=bhat,ci.beta, `exp(beta)` = or, ci.or) |> round(digits = 2)
```
Identify these estimates with the corresponding
values in Table.1(c), and express the odds and the probabilities for each of the categories as functions of the β-parameters
odds of often = e^(beta_0)
odds of notoften = e^(beta_0 + beta_1)
odds of no = e^(beta_0 + beta_2)

p_i (low plasma probability) = e^x_iB / 1 + e^x_iB

Use the regression model to calculate the linear predictor, the odds, and the probability of having a low plasma β-carotene concentration, together with their respective 95 % confidence intervals, for each of the three vitamin use categories. 
Compare the result with the
probabilities in Table.1(c)
```{r}
vituse_pred <- df %>%
  select(vituse, betaplasma, lowplasma_01, lowplasma_hl)
vituse_pred

vituse_pred |> mutate(phat = predict(vituse_glm, type = "response")) -> vituse_pred
vituse_glm$coefficients[1]
odds_0 = exp(vituse_glm$coefficients[1])
odds_1 = exp(vituse_glm$coefficients[1] + vituse_glm$coefficients[2])
odds_2 = exp(vituse_glm$coefficients[1] + vituse_glm$coefficients[3])


p_1 = exp(vituse_glm$coefficients[1]) / (1 + exp(vituse_glm$coefficients[1]))
p_1

p_2 = exp(vituse_glm$coefficients[1] + vituse_glm$coefficients[2]) / (1 + exp(vituse_glm$coefficients[1] + vituse_glm$coefficients[2]))
p_2

p_3 = exp(vituse_glm$coefficients[1] + vituse_glm$coefficients[3]) / (1 + exp(vituse_glm$coefficients[1] + vituse_glm$coefficients[3]))
p_3


cbind("vituse"=c(1,2,3),"odds" = c(odds_0, odds_1, odds_2), "prob_low" = c(p_1,p_2,p_3))

vituse_pred <- data.frame(vituse = c("often", "notoften", "no"))
vituse_pred
lambda <- qnorm(1 - 0.05/2)
vituse_pred <- cbind(vituse_pred,
      logit = predict(vituse_glm, vituse_pred, se.fit = TRUE))
glimpse(vituse_pred)

vituse_pred |> mutate(
  logit.lwr = logit.fit - lambda*logit.se.fit,
  logit.upr = logit.fit + lambda*logit.se.fit) -> vituse_pred
glimpse(vituse_pred)
vituse_pred |> mutate(
  "odds" = c(odds_0, odds_1, odds_2),
  odds.lwr = exp(logit.lwr),
  odds.upr = exp(logit.upr),
  "prob_low" = c(p_1,p_2,p_3)) -> vituse_pred

vituse_pred |> mutate(
  p.lwr = odds.lwr/(1 + odds.lwr),
  p.upr = odds.upr/(1 + odds.upr)) -> vituse_pred
glimpse(vituse_pred)


# Try more effective code
vituse_pred <- predict(vituse_glm, vituse_pred, se.fit = TRUE)
vituse_pred
vituse_x0 <- data.frame(vituse = c("often", "notoften", "no"))

vituse_x0 |> mutate(
  logit = vituse_pred$fit,
  logit.lwr = vituse_pred$fit - lambda*vituse_pred$se.fit,
  logit.upr = vituse_pred$fit + lambda*vituse_pred$se.fit,
  
  "odds" = exp(vituse_pred$fit),
  odds.lwr = exp(logit.lwr),
  odds.upr = exp(logit.upr),
  
  "prob_low" = exp(vituse_pred$fit)/ (1+exp(vituse_pred$fit)),
  p.lwr = odds.lwr/(1 + odds.lwr),
  p.upr = odds.upr/(1 + odds.upr)) -> vituse_x0

vituse_x0
```

## (e)
For Model.1(d), use a suitable test to determine whether there are any significant differences between the vitamin use categories in the model.

Report what type of test you use, the null hypothesis H0, the test statistic, the asymptotic distribution of the test statistic when H0 is true, the P-value and the conclusion.
H_0 : B1=B2
LR-test againsts the null model
calculating D_0 - D and the difference in the number of parameters, i.e., the difference in the degrees of freedom. Calculate the P-value using the Chi-squared distribution

```{r}
D_diff <- vituse_sum$null.deviance - vituse_sum$deviance
df_diff <- vituse_sum$df.null - vituse_sum$df.residual
chi2_alpha <- qchisq(p = 1 - 0.05, df = df_diff)
Pvalue <- pchisq(q = D_diff, df = df_diff, lower.tail = FALSE)

cbind(D_diff, df_diff, chi2_alpha, Pvalue)
```
p-value = 1.811218e-05
Reject H_0, all parameters seem to be significant

# 2 Low plasma β-carotene and BMI